scRNAseq benchmark in want of a title

### 2019-11-27 bmc_article.tex
Updates from Overleaf

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets). Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). Each step can include various parameters, all of which must be given alternative values. The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed. In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure1_doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets. DoubletFinder failed due to an error on the mixology10x3cl dataset. \textbf{C:} Running time of the different methods. \textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e. assigned a wrong cluster. The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
    \label{fig:figure3}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure2_filtering_files/figure-html/figure2-1.png}}
    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
    \label{fig:figure4}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure3_normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
    \label{fig:figure5}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance. \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
    \label{fig:figure6}
\end{figure}

@@

We next investigated the impact of the number of dimensions used on the clustering. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.

\subsubsection*{Rules for selecting components}

@@

All analyses were performed through the \href{pipComp}{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here. All code to reproduce each figure is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

@@

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & source & protocol & nbCells & subpops & description \\ 
  \hline
Koh & GSE85066 & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
  Kumar & GSE60749 & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & \citep{duoClustering2018} & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & \citep{duoClustering2018} & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & \citep{duoClustering2018} & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & \cite{tianMixology2018} & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & \cite{tianMixology2018} & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
   \hline
\end{tabular}
\end{table}

>>>>

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets). In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF]. 
Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \textit{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \textit{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets. DoubletFinder failed due to an error on the mixology10x3cl dataset. \textbf{C:} Running time of the different methods. \textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e. assigned a wrong cluster. The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
    \label{fig:figure3}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
    \label{fig:figure4}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures. A:} Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard Seurat normalization (left) or sctransform (right). The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

@@

\subsubsection*{Estimating dimension}

We next investigated the impact of the number of dimensions used on the separability of the subpopulations. Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact. We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA). As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition. We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields. Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method. Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization. Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e. maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and various dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of dimensionality estimators, using sctransform or standard seurat normalization. \textbf{D:} Clustering accuracy of across the normalization/dimensionality methods.}
    \label{fig:figure8}
\end{figure}

@@

All analyses were performed through the \href{pipeComp}{https://github.com/plger/pipeComp} R package, which implements the pipeline framework described here. All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

@@

\subsection*{Simulated datasets}
The `simMix1` dataset was based on the human PBMC CITE-seq data deposited under `GSE100866`. Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat. We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant such subpopulations. We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations. The `simMix2` dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion. The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & source & protocol & description \\ 
  \hline
Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
   \hline
\end{tabular}
\end{table}
